---
title: "General helper functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General helper functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  dpi = 96
)
```

This article demonstrates the general helper functions in `ibis.insights`. These
helpers are useful for preparing suitability layers, aligning temporal inputs,
and interpreting derivative predictor terms.

```{r libraries, message = FALSE, warning = FALSE}
library(ibis.insights)
library(terra)
library(stars)
```

## Example data

The examples use the small rasters shipped with the package. The land-use layer
is converted from an integer fraction in `[0, 10000]` to a proportion in
`[0, 1]`.

```{r load-data}
range <- terra::rast(system.file(
  "extdata/example_range.tif",
  package = "ibis.insights",
  mustWork = TRUE
))

grassland <- terra::rast(system.file(
  "extdata/Grassland.tif",
  package = "ibis.insights",
  mustWork = TRUE
)) / 10000
names(grassland) <- "grassland"

dem <- terra::rast(system.file(
  "extdata/DEM.tif",
  package = "ibis.insights",
  mustWork = TRUE
))
names(dem) <- "scaled_dem"
```

## Create elevation masks with `create_elevation_mask()`

Use `create_elevation_mask()` to convert a DEM into an elevation suitability
mask from a species-specific preferred interval. A binary mask assigns 1 to
cells inside the interval and 0 outside it. Soft cutoffs keep the interval at 1
and gradually reduce suitability outside the interval to represent uncertainty
around the lower and upper elevation thresholds.

The bundled DEM is scaled to values of 0 and 1. To make the three cutoff
behaviours visible, this example treats values close to 1 as preferred while
values near 0 fall outside the preference range.

```{r elevation-mask-calc}
preferred_elevation <- c(0.85, 1)
elevation_tolerance <- 1

binary_elevation_mask <- create_elevation_mask(
  dem,
  elevation_range = preferred_elevation
)
linear_elevation_mask <- create_elevation_mask(
  dem,
  elevation_range = preferred_elevation,
  cutoff = "linear",
  tolerance = elevation_tolerance
)
exponential_elevation_mask <- create_elevation_mask(
  dem,
  elevation_range = preferred_elevation,
  cutoff = "negative_exponential",
  tolerance = elevation_tolerance
)

names(binary_elevation_mask) <- "binary"
names(linear_elevation_mask) <- "linear"
names(exponential_elevation_mask) <- "negative_exponential"
```

```{r elevation-mask-summary}
preferred_elevation
terra::global(
  c(binary_elevation_mask, linear_elevation_mask, exponential_elevation_mask),
  "range",
  na.rm = TRUE
)
```

```{r elevation-mask-plot}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 3, 4))
plot(binary_elevation_mask, main = "Binary")
plot(linear_elevation_mask, main = "Linear cutoff")
plot(exponential_elevation_mask, main = "Negative exponential")
par(op)
```

The same function also works with `stars` objects.

```{r elevation-mask-stars}
create_elevation_mask(
  stars::st_as_stars(dem),
  elevation_range = preferred_elevation
)
```

## Clamp values with `st_clamp()`

Use `st_clamp()` when a suitability or fractional layer may contain values
outside its valid range. This can happen after rescaling, combining predictors,
or applying a model transformation. Values below the lower bound are set to the
lower bound; values above the upper bound are set to the upper bound.

```{r clamp-calc}
raw_suitability <- (grassland * 1.4) - 0.2
names(raw_suitability) <- "raw_suitability"

clamped_suitability <- st_clamp(
  raw_suitability,
  lower = 0,
  upper = 1
)
names(clamped_suitability) <- "clamped_suitability"
```

```{r clamp-summary}
terra::global(c(raw_suitability, clamped_suitability), "range", na.rm = TRUE)
```

```{r clamp-plot}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 3, 4))
plot(raw_suitability, main = "Before clamping")
plot(clamped_suitability, main = "Clamped to [0, 1]")
par(op)
```

The same function also works with `stars` objects.

```{r clamp-stars}
clamped_stars <- st_clamp(
  stars::st_as_stars(raw_suitability),
  lower = 0,
  upper = 1
)
clamped_stars
```

## Align temporal layers with `align_temporal()`

Use `align_temporal()` when two temporal raster objects use different time
steps. For each target year, the function selects the most recent source layer
whose year is less than or equal to the target year. If the target is earlier
than the first source year, the first source layer is used.

Here a coarse source time series is aligned to more frequent target years.

```{r align-calc}
source_lu <- c(grassland * 0.5, grassland * 0.8, grassland)
names(source_lu) <- c("grassland_2020", "grassland_2040", "grassland_2060")
terra::time(source_lu) <- as.Date(c(
  "2020-01-01",
  "2040-01-01",
  "2060-01-01"
))

target_template <- c(range, range, range, range)
names(target_template) <- c("target_2015", "target_2025", "target_2045", "target_2075")
terra::time(target_template) <- as.Date(c(
  "2015-01-01",
  "2025-01-01",
  "2045-01-01",
  "2075-01-01"
))

aligned_lu <- align_temporal(source_lu, target_template)
```

```{r align-mapping}
data.frame(
  target_year = as.integer(terra::time(aligned_lu)),
  selected_source_layer = names(aligned_lu)
)
```

```{r align-plot-source}
plot(
  source_lu,
  main = c("Source 2020", "Source 2040", "Source 2060")
)
```

```{r align-plot-output}
plot(
  aligned_lu,
  main = c("Target 2015", "Target 2025", "Target 2045", "Target 2075")
)
```

## Recreate derivative predictor ranges with `create_derivate_range()`

`create_derivate_range()` is a specialized helper for reconstructing the range
of an original predictor that is represented by derivative model terms such as
`bin`, `thresh`, or `hinge` features. A typical use is to inspect which values of
an environmental variable are associated with positive derivative coefficients.

The helper expects derivative feature names. In this simple example, a
synthetic temperature predictor is created from the grassland fraction so it has
nonzero values across a useful range. Positive `bin_temperature_*` coefficients
mark values between 4 and 22 as favourable, while a negative coefficient for
higher values is ignored.

```{r derivate-data}
temperature <- grassland * 30
names(temperature) <- "temperature"

coefs <- data.frame(
  Feature = c(
    "(Intercept)",
    "bin_temperature_4_12",
    "bin_temperature_12_22",
    "bin_temperature_22_30"
  ),
  Beta = c(0, 1.2, 0.8, -0.5)
)

temperature_range <- ibis.insights:::create_derivate_range(
  env = temperature,
  varname = "temperature",
  co = coefs,
  to_binary = FALSE
)

temperature_mask <- ibis.insights:::create_derivate_range(
  env = temperature,
  varname = "temperature",
  co = coefs,
  to_binary = TRUE
)
names(temperature_range) <- "temperature_range"
names(temperature_mask) <- "temperature_mask"
```

```{r derivate-summary}
coefs
terra::global(c(temperature, temperature_range, temperature_mask), "range", na.rm = TRUE)
```

```{r derivate-plot}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 3, 4))
plot(temperature, main = "Original predictor")
plot(temperature_range, main = "Positive derivative range")
plot(temperature_mask, main = "Binary range mask")
par(op)
```
