---
title: "InSiGHTS functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{InSiGHTS functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  dpi = 96
)
```

This article gives a first tour of the main `insights_*()` functions. The
examples use the small rasters shipped with the package so they can be rendered
by pkgdown without external downloads.

| Function | Use it when | Main output |
| --- | --- | --- |
| `insights_fraction()` | Land-use, habitat, or condition layers are fractions or suitability weights in `[0, 1]`. | A refined suitability or area-of-habitat raster. |
| `insights_area()` | Land-use or habitat layers are already expressed as area per cell. | A refined raster already in area units. |
| `insights_discount()` | A habitat class should be discounted by an age or maturity layer. | A maturity-weighted version of the land-use layer. |
| `insights_summary()` | Refined rasters need to be converted to totals or relative indicators. | A data frame of absolute area, relative change, or symmetric relative change. |

```{r libraries, message = FALSE, warning = FALSE}
library(ibis.insights)
library(terra)
library(stars)
```

## Example inputs

The current range example is a binary range layer. The land-use layers are stored
as integer fractions from 0 to 10000, so we divide them by 10000 before passing
them to `insights_fraction()`. The DEM example is already scaled to `[0, 1]` and
is used here as a simple additional suitability mask.

```{r load-data}
range <- terra::rast(system.file(
  "extdata/example_range.tif",
  package = "ibis.insights",
  mustWork = TRUE
))

grassland <- terra::rast(system.file(
  "extdata/Grassland.tif",
  package = "ibis.insights",
  mustWork = TRUE
)) / 10000

sparse_vegetation <- terra::rast(system.file(
  "extdata/Sparsely.vegetated.areas.tif",
  package = "ibis.insights",
  mustWork = TRUE
)) / 10000

lu_fraction <- c(grassland, sparse_vegetation)
names(lu_fraction) <- c("grassland", "sparse_vegetation")

extra_suitability <- terra::rast(system.file(
  "extdata/DEM.tif",
  package = "ibis.insights",
  mustWork = TRUE
))
names(extra_suitability) <- "extra_suitability"
```

```{r plot-inputs}
op <- par(mfrow = c(2, 2), mar = c(2, 2, 3, 4))
plot(range, main = "Range")
plot(lu_fraction[[1]], main = "Grassland fraction")
plot(lu_fraction[[2]], main = "Sparse vegetation fraction")
plot(extra_suitability, main = "Additional suitability mask")
par(op)
```

## Fractional land-use: `insights_fraction()`

Use `insights_fraction()` when the habitat or land-use input is a fraction,
probability, or suitability weight. Multiple land-use layers are summed before
they are applied to the range. `clamp = TRUE` caps the combined suitability at
one when several suitable land-use classes overlap.

```{r fraction-calc}
aoh_fraction <- insights_fraction(
  range = range,
  lu = lu_fraction,
  other = extra_suitability,
  clamp = TRUE
)
```

```{r fraction-plot}
plot(aoh_fraction, main = "Fractional InSiGHTS output")
```

`insights_summary()` converts this fractional raster to area by multiplying by
cell area, because `toArea = TRUE` by default.

```{r fraction-summary}
insights_summary(aoh_fraction, relative = FALSE)
```

## Areal land-use: `insights_area()`

Use `insights_area()` when the land-use values are already in area units. In this
example we convert the fractional land-use layers to square kilometres per cell
with `terra::cellSize()`. Because the output is already in area units, summarize
with `toArea = FALSE`.

```{r area-calc}
cell_area_km2 <- terra::cellSize(lu_fraction[[1]], unit = "km")
lu_area_km2 <- lu_fraction * cell_area_km2

aoh_area <- insights_area(
  range = range,
  lu = lu_area_km2,
  other = extra_suitability
)
```

```{r area-plot}
plot(aoh_area, main = "Areal InSiGHTS output (km2 per cell)")
```

```{r area-summary}
insights_summary(aoh_area, toArea = FALSE, relative = FALSE)
```

## Maturity weighting: `insights_discount()`

Use `insights_discount()` before `insights_fraction()` or `insights_area()` when a
habitat class should not count at full value immediately after it appears. The
example below creates a simple dummy age layer for grassland. Real analyses
should use an age or maturity layer for the same habitat class.

```{r discount-calc}
grassland_age <- lu_fraction[["grassland"]] * 40
names(grassland_age) <- "grassland_age_years"

invisible(utils::capture.output({
  grassland_discounted <- insights_discount(
    lu = lu_fraction[["grassland"]],
    age = grassland_age,
    target_age = 20,
    target = 0.95
  )
}))
names(grassland_discounted) <- "grassland_discounted"

lu_discounted <- c(grassland_discounted, lu_fraction[["sparse_vegetation"]])
names(lu_discounted) <- c("grassland_discounted", "sparse_vegetation")

aoh_discounted <- insights_fraction(
  range = range,
  lu = lu_discounted,
  other = extra_suitability,
  clamp = TRUE
)
```

```{r discount-plot}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 3, 4))
plot(lu_fraction[["grassland"]], main = "Original grassland")
plot(grassland_age, main = "Dummy age layer")
plot(grassland_discounted, main = "Discounted grassland")
par(op)
```

```{r discounted-aoh-plot}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 3, 4))
plot(aoh_fraction, main = "Without discount")
plot(aoh_discounted, main = "With grassland discount")
par(op)
```

```{r discounted-summary}
insights_summary(aoh_discounted, relative = FALSE)
```

## Indicators: `insights_summary()`

With a single-layer raster, `insights_summary()` reports an absolute total. With
multiple layers or time steps, it can also report change relative to the first
time step. The standard relative change is returned as a percentage; the
symmetric relative change is bounded in `[-1, 1]`.

```{r summary-series}
aoh_series <- c(aoh_fraction, aoh_discounted, aoh_discounted * 0.8)
names(aoh_series) <- c("baseline", "maturity_discount", "future_example")
terra::time(aoh_series, tstep = "years") <- c(2020, 2040, 2060)

index_summary <- insights_summary(
  aoh_series,
  relative = TRUE,
  symmetric = TRUE
)
index_summary
```

```{r summary-plot}
plot(
  index_summary$time,
  index_summary$relative_change_perc,
  type = "b",
  pch = 19,
  xlab = "Year",
  ylab = "Relative change (%)",
  main = "InSiGHTS index relative to 2020"
)
abline(h = 0, lty = 2, col = "grey50")
```

## Temporal scenario example

The package also ships a small temporal scenario for *Bombina bombina*. This is a
`stars` object with a time dimension and an `ensemble_threshold` attribute. It
is useful for demonstrating how the `insights_*()` functions handle temporal
range inputs.

```{r temporal-data}
bombina_range <- stars::read_mdim(system.file(
  "extdata/Bombina_bombina__ssp126.nc",
  package = "ibis.insights",
  mustWork = TRUE
))
bombina_range <- split(bombina_range)
bombina_range <- bombina_range["ensemble_threshold"]
```

```{r temporal-calc}
invisible(utils::capture.output({
  bombina_aoh <- suppressMessages(insights_fraction(
    range = bombina_range,
    lu = lu_fraction,
    clamp = TRUE
  ))
}))

bombina_summary <- insights_summary(
  bombina_aoh,
  relative = TRUE,
  symmetric = TRUE
)
bombina_summary
```

```{r temporal-plot}
bombina_aoh_raster <- terra::rast(bombina_aoh)
terra::time(bombina_aoh_raster) <- stars::st_get_dimension_values(
  bombina_aoh,
  "time"
)

op <- par(mfrow = c(1, 3), mar = c(2, 2, 3, 4))
plot(bombina_aoh_raster[[1]], main = "2015")
plot(bombina_aoh_raster[[5]], main = "2055")
plot(bombina_aoh_raster[[9]], main = "2095")
par(op)
```

```{r temporal-index-plot}
plot(
  as.Date(bombina_summary$time),
  bombina_summary$relative_change_perc,
  type = "b",
  pch = 19,
  xlab = "Year",
  ylab = "Relative change (%)",
  main = "Temporal InSiGHTS index"
)
abline(h = 0, lty = 2, col = "grey50")
```
